Combinatorics of the geometry of Wilson loop diagrams I: equivalence classes via matroids and polytopes

### 2019-06-28 GWLDCI.tex
susama's changes

\begin{proof}
One direction has been proved in \cite[Theorem 1.18]{wilsonloop}, but we give a different proof here to be consistent with the method of this document.

For the if direction, it suffices to prove the result for two equivalent diagrams that differ in exactly one exact subdiagram. Let $W$ and $W'$ be equivalent Wilson loop diagrams which can be decomposed as
\[W = (P \sqcup R, [n]), \qquad W' = (P \sqcup R', [n]),\]
where $(R,V(R))$ and $(R',V(R'))$ are maximal exact subdiagrams of $W$, $W'$ respectively, and $V(R) = V(R')$. By Theorem \ref{exactuniformthm}, $(R, V(R))$ defines a uniform matroid. Therefore any set $U \subseteq V(R)$ of size $|R|$ is independent in the subdiagram $(R, V(R))$, and thus in $W$.

Now consider the complement $V(R)^c$, which is just the propagator flat $F(R^c)$. By Lemma \ref{maxexactcomplementrank}, $F(R^c)$ is a cyclic flat of rank $|R^c| = |P|$ in $M(W)$. Let $B \subseteq V(R)^c$ be a maximal independent subset of $V(R)^c$, i.e. $\rk (B) = |P|$. Since $V(R)^c$ is a flat, adding any element of $V(R)$ to $B$ increases the rank of the set.

Therefore, any basis of $W$ can be written as $B \sqcup U$ for some $B$ and $U$ of the form indicated described above. We may apply the same arguments to $W'$.

However, since $F(P)$ is common to both $W$ and $W'$, and $V(R) = V(R')$, any basis of $W'$ can also be written $B \cup U$. Thus both matroids have the same bases sets, proving that they are the same.

For the converse, assume that the matroids associated to $W$ and $W'$ are the same: $M(W) = M(W')= M$. Let $\mathcal{R} = \{(P_i, V(P_i)\}_{i=1}^m$ and $\mathcal{R}' = \{(P'_i, V(P'_i)\}_{i=1}^l$ be the sets of maximal exact subdiagrams of $W$ and $W'$. Recall that these sets $\{P_i\}$ and $\{P_i'\}$ partition the propagator sets $\cP$ and $\cP'$ of $W$ and $W'$ respectivley. Let $\mathcal{F}$ and $\mathcal{F}'$ the respective complementary vertices: \bas \mathcal{F} = \{F_i = V(P_i)^c | (P_i, V(P_i)) \in \mathcal{R}\} \quad  \mathcal{F}' = \{F_i' = V(P'_i)^c | (P'_i, V(P'_i)) \in \mathcal{R}\} \;.\eas By Theorem \ref{maxexactcomplementrank}, every $F_i$ and $F_i'$ is a cyclic flat in $M$. They are also propagator flats in $W$ and $W'$ of rank equal to the number of propagators defining it. We claim that $\mathcal{F} = \mathcal{F'}$.

If not, there would be a cyclic flat $F_j$ of $M$ in one set but not the other. Without loss of generality, write $F_j \in \mathcal{F} \setminus \mathcal{F}'$.  We may write $F_j = V(P_j)^c$, with $(P_j, V(P_j))$ a maximal exact subdiagram in $W$. By Lemma \ref{lem facts about WLD matroids} (5), we may write $F_j= F(Q)$ as a propagator flat. Therefore, Lemma \ref{maxexactcomplementrank}, $\rk(F_j) = |Q|$. By Lemma \ref{contractsubdiaglem}, $M/F(Q) = F/F_j = M(Q^c, V(Q^c))$, which is uniform, by construction of $F_j$. Theorem \ref{exactuniformthm} $(Q^c, V(Q^c))$ is exact. Since this exact subdiagram is not in $\mathcal{R}'$ by construction of $F_j$, it is not a maximal exact subdiagram of $W'$. That is, there is a maximal exact subdiagram containing it: $(Q^c, V(Q^c)) \subsetneq (P_j', V(P_j'))$. Thus there is a $F_j' \in \mathcal{R}'$ such that $F_j' \subsetneq F_j$. We show that $F_j$ is not a cyclic flat.

Suppose the exact diagram  $(P_j', V(P_j'))$ has $m$ more propagators than $(Q^c, V(Q^c))$: $|P_j' \setminus Q^c| = m$. Since $(P_j', V(P_j'))$ defines a uniform matroid, any subset of vertices numbering fewer than the number of propagators in $P_j'$ forms an independent flat of $M(P_j', V(P_j'))$, and thus an independent set of $M$. \sanote{is it worth citing a result that says for $W \subset W'$, if $V$ independent in $M(W)$ then also independent in $M(W')$?} Next, compare the number of vertices in $F_j'$ and $F_j$ to see that $F_j'$ has $m$ more vertices than $F_j$:  \bas |F_j| = n -|Q^c| -3  \textrm{ while } |F_j'| = n -|P_j'| -3 = n -|Q^c| -m -3\;.\eas So $|F_j \setminus F_j' | = m$ and $F_j \setminus F_j' = V(P_j')\setminus V(Q^c)$.  Thus $F_j \setminus F_j'$ is an independent set in $M$. Therefore $F_j$ is the union of a cyclic flat, $F_j'$, and an independent set, $F_j \setminus F_j'$. Therefore, $F_j$ is not a cyclic flat. Therefore, $F_j \not \in \mathcal{F} \setminus \mathcal{F}'$. Therefore, $\mathcal{F} = \mathcal{F}'$, as claimed.

Since $\mathcal{F} = \mathcal{F'}$, the maximally exact subdiagrams of $W$ and $W'$, are supported on the same vertex sets. All propagators of $W$ and $W'$ are accounted for in these sets. There there exists a transitive chain of equivalent diagrams that put $W$ and $W'$ in the same equivalence class.
\end{proof}

\begin{proof}[Proposed alternative proof] \sanote{Looks good to me}
One direction has been proved in \cite[Theorem 1.18]{wilsonloop}, but we give a different proof here to be consistent with the method of this paper.

@@

Now consider the propagator flat $F(P)$, which is a cyclic flat of rank $|P|$ in both $M(W)$ and $M(W')$ by Lemma \ref{maxexactcomplementrank}. Let $B \subseteq F(P)$ be an independent set of rank $|P|$. Adding any element of $V(R)$ to $B$ increases its rank in both matroids since $F(P)$ is a flat and $F(P) \cap V(R) = \emptyset$, so we can lift any independent set in $F(P)$ to an independent set of both $M(W)$ and $M(W')$ by adjoining any independent set in $V(R)$. In particular, $B\sqcup U$ is a basis in both $M(W)$ and $M(W')$ for any $U \subseteq V(R)$ with $|U| = |R|$.

@@

We also need the following notation: given a polygon dissection $D$ of an $n$-gon, let $V_{>3}(D)$ be the set of vertices of the polygon which are on at least one non-triangle face of $D$.  (Note that $D$ may have many non-triangle faces and so $V_{>3}(D)$ will be the union of the vertices of each non-triangle face.) We also write $[x_i,x_j]$ for the set of vertices from $x_i$ to $x_j$ including the endpoints, and $(x_i,x_j)$ for the same set but excluding the endpoints. \sanote{make a dfn?}

@@

\note{I wonder if either of you see a shorter or more conceptual way to prove this.}

@@

As observed in the proof of Proposition~\ref{prop easy way}, \st{the support of any non-zero vector parallel to an edge is on the indices corresponding to the four vertices bounding the quadrilateral of the dissection associated with the edge.}{\color{violet}any non-zero vector in $\R^n$ paralell to an edge of $A_n$ has non-zero entries only in the coordinates corresponding to the four vertices bounding the quadrilateral of the dissection associated with the edge.}  Thus if $V_{>3}(\tau(W)) \neq V_{>3}(\tau(X))$ \hlfix{(notation as defined before Lemma {\ref{lem good quads}})}{refer to dfn?} then the span of the vectors parallel to their edges has different support and so they cannot be parallel.  So we can assume that $V_{>3}(\tau(W)) = V_{>3}(\tau(X))$.

@@

Propositions~\ref{prop easy way} and \ref{prop hard way} together imply the main result of this section:
\begin{thm}\label{thm:count inequiv diagrams}
  The number of inequivalent weakly admissible Wilson loop diagrams on $n$ is the number of non-parallel faces in the associahedron $A_n$.
\end{thm} \sanote{this is the main result of the section}

\note{Do we want a conclusion section?}

\bibliographystyle{abbrv}
\bibliography{WLDbib}

>>>>

\begin{proof}One direction has been proved in \cite[Theorem 1.18]{wilsonloop}, but we give a different proof here to be consistent with the method of this paper.

@@

Now consider the propagator flat $F(P)$, whichf is a cyclic flat of rank $|P|$ in both $M(W)$ and $M(W')$ by Lemma \ref{maxexactcomplementrank}. Let $B \subseteq F(P)$ be an independent set of rank $|P|$. Adding any element of $V(R)$ to $B$ increases its rank in both matroids since $F(P)$ is a flat and $F(P) \cap V(R) = \emptyset$, so we can lift any independent set in $F(P)$ to an independent set of both $M(W)$ and $M(W')$ by adjoining any independent set in $V(R)$. In particular, $B\sqcup U$ is a basis in both $M(W)$ and $M(W')$ for any $U \subseteq V(R)$ with $|U| = |R|$.

@@

We also need the following notation: given a polygon dissection $D$ of an $n$-gon, let $V_{>3}(D)$ be the set of vertices of the polygon which are on at least one non-triangle face of $D$.  (Note that $D$ may have many non-triangle faces and so $V_{>3}(D)$ will be the union of the vertices of each non-triangle face.) We also write $[x_i,x_j]$ for the set of vertices from $x_i$ to $x_j$ including the endpoints, and $(x_i,x_j)$ for the same set but excluding the endpoints.

@@

@@

As observed in the proof of Proposition~\ref{prop easy way}, any non-zero vector in $\R^n$ parallel to an edge of $A_n$ has non-zero entries only in the coordinates corresponding to the four vertices bounding the quadrilateral of the dissection associated with the edge.  Thus if $V_{>3}(\tau(W)) \neq V_{>3}(\tau(X))$ (notation as defined before Lemma \ref{lem good quads}) then the span of the vectors parallel to their edges has different support and so they cannot be parallel.  So we can assume that $V_{>3}(\tau(W)) = V_{>3}(\tau(X))$.

@@

Propositions~\ref{prop easy way} and \ref{prop hard way} together imply the main result of this section:
\begin{thm}\label{thm:count inequiv diagrams}
  The number of inequivalent weakly admissible Wilson loop diagrams on $n$ is the number of non-parallel faces in the associahedron $A_n$.
\end{thm} 
\bibliographystyle{abbrv}
\bibliography{WLDbib}
